Concept note

### 2018-05-06 inst/rmd/externalities.tex
externalities text

To test whether IRS from one household has an effect on workers living
in another household, we first must make some assumptions about the
level and waning effect of IRS generally. For our purposes, we make a
rather simple assumption, based on the literature: that IRS has an
immediate effect on mosquitoes and should affect malaria risk following
the incubation period, that the waning effect begins in month 4 (Tukei
et al., 2017), that IRS effects those not living in the household by
reducing reproduction of mosquitoes (White et al., 2011), and that IRS'
effect after 1 year is essentially null (Bukirwa et al., 2009). We
simplify these assumptions into a time-protection matrix, which can be
visualized in the below chart.

\begin{center}\includegraphics{externalities_files/figure-latex/unnamed-chunk-3-1} \end{center}

The above chart shows two lines: the orange line is based on Tukei's
emperical results regarding the protection afforded by IRS against
malaria infection (among children) as a function of time since IRS; the
blue line is our simplification of Tukei's work.

@@

Having estimated each household's protection level at all times (ie, the
``amount'' of protection that they pass on to neighbors), we need to
define a method for how much of that total level is passed on. A
household which is 1 meter from a neighbor should, theoretically, afford
more protection to that household than another which is 1000 meters. To
account for the effect of distance, we define a simple function for
weighting a nearby residence's contribution to a household's ``herd
protection'' score as follows:

@@

We heavily weight towards closer houses to account for the relatively
small travel distances that most mosquitoes fly in normal conditions
(Verdonschot and Besse-Lototskaya, 2014).

@@

We consider a house's ``herd'' protection level (ie, the protection
conferred to the house through externality) to be the average of the
other houses' non-herd protection levels, weighted by the distance to
the house in question. The below is an illustration of a house (the red
x) and the neighboring houses (circles). The ``weight'' of each house is
indicated by the circle size, and the protection level of the house is
indicated by the shading of the circle.

@@

This score is conceptually similar to Cohen and Dupas' quantification of
the positive externalities of ITN (bednet) use in Kenya (Cohen and
Dupas, 2010) in that it attempts to estimate the protection conferred to
``non-users'' by users and uses a ``weighted sum'' (as opposed to
average). Our approach differs in that the time dimension for IRS is
much more important than ITN (ie, IRS is not a binary but rather a
waning function, as described previously), requiring us to create herd
scores for every day.

Our approach can be justified in that previous studies have found strong
positive health externalities in malaria interventions related to ITN
coverage (Alaii et al., 2003). No studies exist on positive
externalities for IRS coverage, but to the extent that the mechanisms
for the reduction in infection are similar (reduction in the natural
reservoir of the disease, reduction in the number of vectors, etc.), it
is reasonable to assume similar effects. Additionally, we using weighted
distance for our score calculation (rather than simply defining a radius
threshold) because of previous studies' findings that there is a linear
decline in protection conferred to others with greater distance from an
intervention (Binka et al., 2007)

@@

We calculate a ``herd protection'' score for each location-date
combination (1759100 values). Whereas our original model specification
looked like:

@@

\begin{longtable}[t]{ll}
\caption{\label{tab:unnamed-chunk-8}All absenteeism with herd immunity: model results}\\
\toprule
Term & Estimate\\
\midrule
\endfirsthead
\caption[]{All absenteeism with herd immunity: model results \textit{(continued)}}\\
\toprule
Term & Estimate\\
\midrule
\endhead
\
\endfoot
\bottomrule
\endlastfoot
\addlinespace[1.5em]
\multicolumn{2}{l}{\textbf{Permanent field worker}}\\
\hspace{1em}Malaria season & 3.597 (P<0.001)\\
\hspace{1em}Months since IRS: IRS & 6.149 (P<0.001)\\
\hspace{1em}Rainy day & 3.246 (P<0.001)\\
\hspace{1em}Herd protection & -0.01 (P<0.001)\\
\hspace{1em}Malaria season:Months since IRS: IRS & -8.548 (P<0.001)\\
\addlinespace[1.5em]
\multicolumn{2}{l}{\textbf{Permanent not field worker}}\\
\hspace{1em}Malaria season & 0.792 (P<0.001)\\
\hspace{1em}Months since IRS: IRS & 2.243 (P<0.001)\\
\hspace{1em}Rainy day & 4.552 (P<0.001)\\
\hspace{1em}Herd protection & -0.009 (P<0.001)\\
\hspace{1em}Malaria season:Months since IRS: IRS & -1.508 (P=0.002)\\
\addlinespace[1.5em]
\multicolumn{2}{l}{\textbf{Temporary field worker}}\\
\hspace{1em}Malaria season & -0.362 (P<0.001)\\
\hspace{1em}Months since IRS: IRS & -0.229 (P=0.004)\\
\hspace{1em}Rainy day & 0.037 (P=0.261)\\
\hspace{1em}Herd protection & 0.002 (P<0.001)\\
\hspace{1em}Malaria season:Months since IRS: IRS & 0.26 (P=0.014)\\
\addlinespace[1.5em]
\multicolumn{2}{l}{\textbf{Temporary not field worker}}\\
\hspace{1em}Malaria season & 3.151 (P<0.001)\\
\hspace{1em}Months since IRS: IRS & -1.718 (P=0.076)\\
\hspace{1em}Rainy day & 0.224 (P=0.469)\\
\hspace{1em}Herd protection & 0.002 (P=0.772)\\
\hspace{1em}Malaria season:Months since IRS: IRS & 2.657 (P=0.02)\\*
\end{longtable}

\subsubsection{HIV}\label{hiv}

We incorporated HIV prevalence of the area lived in, but it had not
significance and was then removed from the model.

@@

@@

The below uses kernel density interpolation (via krieging) to estimate a
generalized protection surface over the entire facility.

@@

\begin{center}\includegraphics{externalities_files/figure-latex/unnamed-chunk-12-1} \end{center}

@@

@@

\begin{center}\includegraphics{externalities_files/figure-latex/unnamed-chunk-16-1} \end{center}

\begin{itemize}
\tightlist
\item
  How to translate the effect into costs and savings (simulation based
  approach at various IRS coverage levels?).
\item
  How to combine this effect with the other potential spillover effect
  (ie, core-periphery issues).
\item
  How / where to report on this in the paper (is this a ``method'' or
  ``result''?).
\end{itemize}

@@

\hypertarget{refs}{}
\hypertarget{ref-Alaii2003FactorsAU}{}
Alaii, J.A., Hawley, W.A., Kolczak, M.S., Kuile, F.O.T., Gimnig, J.E.,
Vulule, J., Odhacha, A., Oloo, A.J., Nahlen, B.L., Phillips-Howard,
P.A., 2003. Factors affecting use of permethrin-treated bed nets during
a randomized controlled trial in western kenya. The American journal of
tropical medicine and hygiene 68 4 Suppl, 137--41.

\hypertarget{ref-Binka2007}{}
Binka, F.N., Kubaje, A., Adjuik, M., Williams, L.A., Lengeler, C.,
Maude, G.H., Armah, G.E., Kajihara, B., Adiamah, J.H., Smith, P.G.,
2007. Impact of permethrin impregnated bednets on child mortality in
kassena-nankana district, ghana: A randomized controlled trial. Tropical
Medicine \& International Health 1, 147--154.
\url{https://doi.org/10.1111/j.1365-3156.1996.tb00020.x}

\hypertarget{ref-Bukirwa2009}{}
Bukirwa, H., Yau, V., Kigozi, R., Filler, S., Quick, L., Lugemwa, M.,
Dissanayake, G., Kamya, M., Wabwire-Mangen, F., Dorsey, G., 2009.
Assessing the impact of indoor residual spraying on malaria morbidity
using a sentinel site surveillance system in western uganda. American
Journal of Tropical Medicine and Hygiene 81, 611--614.
\url{https://doi.org/10.4269/ajtmh.2009.09-0126}

\hypertarget{ref-Cohen2010}{}
Cohen, J., Dupas, P., 2010. Free distribution or cost-sharing? Evidence
from a randomized malaria prevention experiment. Quarterly Journal of
Economics 125, 1--45. \url{https://doi.org/10.1162/qjec.2010.125.1.1}

\hypertarget{ref-Tukei2017}{}
Tukei, B.B., Beke, A., Lamadrid-Figueroa, H., 2017. Assessing the effect
of indoor residual spraying (IRS) on malaria morbidity in northern
uganda: A before and after study. Malaria Journal 16.
\url{https://doi.org/10.1186/s12936-016-1652-4}

\hypertarget{ref-Verdonschot2014}{}
Verdonschot, P.F., Besse-Lototskaya, A.A., 2014. Flight distance of
mosquitoes (culicidae): A metadata analysis to support the management of
barrier zones around rewetted and newly constructed wetlands.
Limnologica - Ecology and Management of Inland Waters 45, 69--79.
\url{https://doi.org/10.1016/j.limno.2013.11.002}

\hypertarget{ref-White2011}{}
White, M.T., Griffin, J.T., Churcher, T.S., Ferguson, N.M., Basáñez,
M.-G., Ghani, A.C., 2011. Modelling the impact of vector control
interventions on anopheles gambiae population dynamics. Parasites \&
Vectors 4, 153. \url{https://doi.org/10.1186/1756-3305-4-153}

>>>>

To test whether IRS from one household has an effect on workers living
in another household, we first must make some assumptions about the
level and waning effect of IRS generally. It is known from the
literature that IRS has an immediate effect on mosquitoes and should
affect malaria risk following the incubation period, that the waning
effect begins in month 4 (Tukei, Beke, and Lamadrid-Figueroa 2017), that
IRS effects those not living in the household by reducing reproduction
of mosquitoes (White et al. 2011), and that IRS' effect after 1 year is
essentially null (Bukirwa et al. 2009).

For our purposes, we make simplify even further. We make no assumptions
about the level of positive externalities afforded by IRS (instead
preferring to estimate this within our model), and we do not account for
waning, instead modeling the entire ``post'' period of 6 months as
one.\footnote{Though not as granular as estimating time-specific protection within the 6 months following IRS administration, our approach is not less accurate; it simply aggregates the entirety of the 6 months after IRS administration into one value, and assumes (as is suggested by the literature) that the period after 6 months is effectively equivalent in chemical terms as the period prior to IRS.}

@@

At any given time, each household is either (a) protected (ie,
\textless{} 6 months following IRS administration) or (b) unprotected.
If the former, they confer protection to their neighbors; if the latter,
they confer no protection to their neighbors. Having assigned each
household's protection status at all times, we need to define a method
for how much of any given household's protection level is passed on. A
household which is 1 meter from a neighbor should, theoretically, afford
more protection to that household than another which is 1000 meters. To
account for the effect of distance, we define a simple function for
weighting a nearby residence's contribution to a household's ``herd
protection'' score as 1 divided by the distance, as follows:

@@

We heavily weight towards closer houses to account for the relatively
small travel distances that most mosquitoes fly in normal conditions
(Verdonschot and Besse-Lototskaya 2014).

@@

We consider a house's ``herd'' protection level (ie, the protection
conferred to the house through externality) to be the average of the
other houses' non-herd protection levels (ie, 0 if not protected and 1
if protected), weighted by the distance to the house in question. The
below is an illustration of a house (the red x) and the neighboring
houses (circles). The ``weight'' of each house is indicated by the
circle size, and the protection level of the house is indicated by the
shading of the circle.

@@

This score is conceptually similar to Cohen and Dupas' quantification of
the positive externalities of ITN (bednet) use in Kenya (Cohen and Dupas
2010) in that it attempts to estimate the protection conferred to
``non-users'' by users and uses a ``weighted sum'' (as opposed to
average). Our approach differs in that the time dimension for IRS is
much more important than ITN (ie, IRS is not a binary but rather a
waning function, as described previously), requiring us to create herd
scores for every day. This is because the protection conferred to a
house by its neighbors may change from one day to the next, when a
neighbor either (a) gets sprayed or (b) leaves ``protected'' status (ie,
due to the end of the 6 months window).

Our approach can be justified in that previous studies have found strong
positive health externalities in malaria interventions related to ITN
coverage (Alaii et al. 2003). No studies exist on positive externalities
for IRS coverage, but to the extent that the mechanisms for the
reduction in infection are similar (reduction in the natural reservoir
of the disease, reduction in the number of vectors, etc.), it is
reasonable to assume similar effects. Additionally, we using weighted
distance for our score calculation (rather than simply defining a radius
threshold) because of previous studies' findings that there is a linear
decline in protection conferred to others with greater distance from an
intervention (Binka et al. 2007)

@@

We calculate a ``herd protection'' score for each location-date
combination (395019 values). Whereas our original model specification
looked like:

@@

\begin{longtable}[t]{ll}
\caption{\label{tab:unnamed-chunk-8}All absenteeism with herd immunity: model results}\\
\toprule
Term & Estimate\\
\midrule
\endfirsthead
\caption[]{All absenteeism with herd immunity: model results \textit{(continued)}}\\
\toprule
Term & Estimate\\
\midrule
\endhead
\
\endfoot
\bottomrule
\endlastfoot
\addlinespace[1.5em]
\multicolumn{2}{l}{\textbf{Permanent field worker}}\\
\hspace{1em}Malaria season & 2.809 (P<0.001)\\
\hspace{1em}IRS status=After & 1.835 (P<0.001)\\
\hspace{1em}Rainy day & 3.765 (P<0.001)\\
\hspace{1em}Herd protection & -0.003 (P=0.368)\\
\hspace{1em}Malaria season:IRS status=After & -5.106 (P<0.001)\\
\addlinespace[1.5em]
\multicolumn{2}{l}{\textbf{Permanent not field worker}}\\
\hspace{1em}Malaria season & 0.831 (P<0.001)\\
\hspace{1em}IRS status=After & 1.151 (P<0.001)\\
\hspace{1em}Rainy day & 3.835 (P<0.001)\\
\hspace{1em}Herd protection & 0 (P=0.927)\\
\hspace{1em}Malaria season:IRS status=After & -1.66 (P<0.001)\\
\addlinespace[1.5em]
\multicolumn{2}{l}{\textbf{Temporary field worker}}\\
\hspace{1em}Malaria season & -0.188 (P=0.001)\\
\hspace{1em}IRS status=After & -0.254 (P<0.001)\\
\hspace{1em}Rainy day & 0 (P=0.994)\\
\hspace{1em}Herd protection & -0.001 (P=0.006)\\
\hspace{1em}Malaria season:IRS status=After & 0.054 (P=0.568)\\
\addlinespace[1.5em]
\multicolumn{2}{l}{\textbf{Temporary not field worker}}\\
\hspace{1em}Malaria season & 5.772 (P<0.001)\\
\hspace{1em}IRS status=After & -2.187 (P=0.023)\\
\hspace{1em}Rainy day & -1.349 (P=0.003)\\
\hspace{1em}Herd protection & 0.043 (P<0.001)\\
\hspace{1em}Malaria season:IRS status=After & -3.235 (P=0.003)\\*
\end{longtable}

@@

The below uses kernel density interpolation to estimate a generalized
protection surface over the entire facility.

@@

@@

\begin{itemize}
\tightlist
\item
  How to translate the effect into costs and savings (simulation based
  approach at various IRS coverage levels?).
\item
  How / where to report on this in the paper (is this a ``method'' or
  ``result''?).
\end{itemize}

\subsection{Appendix}\label{appendix}

@@

\begin{center}\includegraphics{externalities_files/figure-latex/unnamed-chunk-13-1} \end{center}

@@

@@

\hypertarget{refs}{}
\hypertarget{ref-Alaii2003FactorsAU}{}
Alaii, Jane A., William A. Hawley, Margarette S. Kolczak, Feiko O Ter
Kuile, John E. Gimnig, John Vulule, Amos Odhacha, Aggrey J Oloo, Bernard
L. Nahlen, and Penelope A Phillips-Howard. 2003. ``Factors Affecting Use
of Permethrin-Treated Bed Nets During a Randomized Controlled Trial in
Western Kenya.'' \emph{The American Journal of Tropical Medicine and
Hygiene} 68 4 Suppl: 137--41.

\hypertarget{ref-Binka2007}{}
Binka, F. N., A. Kubaje, M. Adjuik, L. A. Williams, C. Lengeler, G. H.
Maude, G. E. Armah, B. Kajihara, J. H. Adiamah, and P. G. Smith. 2007.
``Impact of Permethrin Impregnated Bednets on Child Mortality in
Kassena-Nankana District, Ghana: A Randomized Controlled Trial.''
\emph{Tropical Medicine \& International Health} 1 (2). Wiley-Blackwell:
147--54.
doi:\href{https://doi.org/10.1111/j.1365-3156.1996.tb00020.x}{10.1111/j.1365-3156.1996.tb00020.x}.

\hypertarget{ref-Bukirwa2009}{}
Bukirwa, H., V. Yau, R. Kigozi, S. Filler, L. Quick, M. Lugemwa, G.
Dissanayake, M. Kamya, F. Wabwire-Mangen, and G. Dorsey. 2009.
``Assessing the Impact of Indoor Residual Spraying on Malaria Morbidity
Using a Sentinel Site Surveillance System in Western Uganda.''
\emph{American Journal of Tropical Medicine and Hygiene} 81 (4).
American Society of Tropical Medicine; Hygiene: 611--14.
doi:\href{https://doi.org/10.4269/ajtmh.2009.09-0126}{10.4269/ajtmh.2009.09-0126}.

\hypertarget{ref-Cohen2010}{}
Cohen, Jessica, and Pascaline Dupas. 2010. ``Free Distribution or
Cost-Sharing? Evidence from a Randomized Malaria Prevention
Experiment.'' \emph{Quarterly Journal of Economics} 125 (1). Oxford
University Press (OUP): 1--45.
doi:\href{https://doi.org/10.1162/qjec.2010.125.1.1}{10.1162/qjec.2010.125.1.1}.

\hypertarget{ref-Tukei2017}{}
Tukei, Betty Bawuba, Andy Beke, and Héctor Lamadrid-Figueroa. 2017.
``Assessing the Effect of Indoor Residual Spraying (IRS) on Malaria
Morbidity in Northern Uganda: A Before and After Study.'' \emph{Malaria
Journal} 16 (1). Springer Nature.
doi:\href{https://doi.org/10.1186/s12936-016-1652-4}{10.1186/s12936-016-1652-4}.

\hypertarget{ref-Verdonschot2014}{}
Verdonschot, Piet F.M., and Anna A. Besse-Lototskaya. 2014. ``Flight
Distance of Mosquitoes (Culicidae): A Metadata Analysis to Support the
Management of Barrier Zones Around Rewetted and Newly Constructed
Wetlands.'' \emph{Limnologica - Ecology and Management of Inland Waters}
45 (March). Elsevier BV: 69--79.
doi:\href{https://doi.org/10.1016/j.limno.2013.11.002}{10.1016/j.limno.2013.11.002}.

\hypertarget{ref-White2011}{}
White, Michael T, Jamie T Griffin, Thomas S Churcher, Neil M Ferguson,
María-Gloria Basáñez, and Azra C Ghani. 2011. ``Modelling the Impact of
Vector Control Interventions on Anopheles Gambiae Population Dynamics.''
\emph{Parasites \& Vectors} 4 (1). Springer Nature: 153.
doi:\href{https://doi.org/10.1186/1756-3305-4-153}{10.1186/1756-3305-4-153}.
